---
title: "**Querying Census Data: Census Blocks**"
author: "Robert J. Dellinger"
date: "2024"
---

------------------------------------------------------------------------

# Methods

## Census Data Collection

This project was developed to enable reproducible, efficient access to detailed census block group=l data for environmental studies. The intent is to support work that examines the human and environmental consequences of structural inequality, particularly through the lens of race, income, housing, and related social indicators.

We obtained demographic, housing, and income data from the 2016–2020 American Community Survey (ACS) 5-Year Estimates at the block group level  using the tidycensus R package (Walker, 2023). Data were accessed via the U.S. Census API, and spatial geometries were downloaded directly using the geometry = TRUE argument, leveraging TIGER/Line shapefiles via the tigris package.

We requested variables from the following ACS tables:
	•	B03002: Race and Hispanic origin
	•	B25001: Total housing units
	•	B25003: Housing tenure (owner vs. renter)
	•	B19013: Median household income

To speed up download and support scale-up to national datasets, we used the furrr package to apply get_acs() across counties in parallel, using all available CPU threads except one.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load the necessary libraries
library(furrr)       # For parallel processing
library(purrr)       # For reduce function
library(tidycensus)  # To access ACS data
library(tigris)      # For FIPS codes and spatial data
library(sf)         # For handling spatial data
library(ggplot2)     # For plotting
```

## Load the necessary libraries and setup options

This chunk loads the required libraries and sets up options for the `tigris` and `tidycensus` packages. It also sets up parallel processing using the `furrr` package.


```{r load-options}

# Set tigris options for spatial data handling
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)

# Set your Census API key (replace with your own key)
#census_api_key("YOUR CENSUS API KEY HERE", install = TRUE)

# Set up parallel processing using all but one of your available cores
num_cores <- availableCores() - 1
print(paste("Thread count to use:", num_cores))
plan(multisession, workers = num_cores)
```

## Download ACS Block Group Data for States

This chunk retrieves race, housing, and income data from ACS for all block groups in California, including geographic shapefiles.

```{r census-block-state, message=FALSE, warning=FALSE}

# select the state of interest here 
fips.ca <- "CA"

# Run with same structure (for consistency with national code)
state_bg_sf <- reduce(
  future_map(fips.ca, function(x) {
    get_acs(
      geography = "block group", 
      variables = c(
        "pop"         = "B03002_001",  # Total population
        "pop_nhwhite" = "B03002_003",  # Non-Hispanic White
        "pop_nhblack" = "B03002_004",  # Non-Hispanic Black
        "popx_nhamind"= "B03002_005",  # Non-Hispanic American Indian
        "popx_nhasian"= "B03002_006",  # Non-Hispanic Asian
        "popx_nhhwnpi"= "B03002_007",  # Non-Hispanic Hawaiian/PI
        "popx_nhother"= "B03002_008",  # Non-Hispanic Other
        "popx_nhtwomr"= "B03002_009",  # Non-Hispanic Two+
        "pop_hispltx" = "B03002_012",  # Hispanic/Latinx
        "hu_total"    = "B25001_001",  # Total housing units
        "hu_totocc"   = "B25003_001",  # Occupied housing units
        "hu_totown"   = "B25003_002",  # Owner-occupied
        "hu_totrnt"   = "B25003_003",  # Renter-occupied
        "hh_total"    = "B19001_001",  # Total households
        "hh_medinc"   = "B19013_001"   # Median household income
      ), 
      year = 2019,
      survey = "acs5",
      state = x,
      geometry = TRUE,
      output = "wide",
      keep_geo_vars = TRUE
    )
  }, .progress = TRUE),
  rbind
)

# Plot map
state_bg_map <- ggplot(data = ca_bg_sf) +
  geom_sf(aes(fill = hh_medincE), color = NA) +
  viridis::scale_fill_viridis(
    name = "Median Income ($)",
    option = "plasma",
    trans = "log",  # Useful for skewed income data
    na.value = "grey90"
  ) +
  theme_minimal() +
  labs(
    title = "Median Household Income by Block Group (California, ACS 2019)",
    caption = "Source: U.S. Census Bureau, 2019 ACS 5-Year Estimates"
  )

Print(state_bg_map)
ggsave(state_ct_map, filename = here::here("Outputs", "Figures", "ca_income_ct_map.png"), 
       width = 10, height = 6, dpi = 300)

```

## Download ACS Data for the Entire United States

This chunk retrieves the same demographic and housing variables as above, but for all 50 states + DC.


```{r census-block-national, message=FALSE, warning=FALSE}


# Get a list of state abbreviations for the 50 states plus DC.
# fips_codes is provided with tidycensus.
fips.states <- unique(fips_codes$state)[1:51]

# Use future_map to download data for each state in parallel and then combine the results
ntl_bg_sf <- reduce(
  future_map(fips.states, function(x) {
    get_acs(
      geography = "block group", 
      variables = c(
        "pop"         = "B03002_001",  # Total population
        "pop_nhwhite" = "B03002_003",  # Non-Hispanic White
        "pop_nhblack" = "B03002_004",  # Non-Hispanic Black
        "popx_nhamind"= "B03002_005",  # Non-Hispanic American Indian/Alaska Native
        "popx_nhasian"= "B03002_006",  # Non-Hispanic Asian
        "popx_nhhwnpi"= "B03002_007",  # Non-Hispanic Hawaiian/Pacific Islander
        "popx_nhother"= "B03002_008",  # Non-Hispanic Other
        "popx_nhtwomr"= "B03002_009",  # Non-Hispanic Two or More Races
        "pop_hispltx" = "B03002_012",  # Hispanic/Latinx
        "hu_total"    = "B25001_001",  # Total housing units
        "hu_totocc"   = "B25003_001",  # Occupied housing units
        "hu_totown"   = "B25003_002",  # Owner-occupied housing units
        "hu_totrnt"   = "B25003_003",  # Renter-occupied housing units
        "hh_total"    = "B19001_001",  # Total number of households
        "hh_medinc"   = "B19013_001"   # Median household income
      ),
      year = 2019,
      survey = "acs5",
      state = x, 
      geometry = TRUE, 
      output = "wide",
      keep_geo_vars = TRUE
    )
  }, .progress = TRUE),  # Displays a progress bar during the parallel execution
  rbind  # Combine the list of state data frames into one large spatial data frame
)

# Optionally, you could save the combined data for future use
#sf::st_write(ntl_bg_sf, here::here("Data", "Processed", "ntl_bg_sf.shp"))

```


## Visualize National Median Income

This final chunk filters out rows with missing income and maps national median income at the block group level.

```{r plotting-national-census-data, message=FALSE, warning=FALSE}


# Filter out geometries with missing income data
ntl_income <- ntl_bg_sf %>%
  filter(!is.na(hh_medincE))

# Plot median household income by block group
ntl_bg_income_map <- ggplot(data = ntl_income) +
  geom_sf(aes(fill = hh_medincE), color = NA) +
  scale_fill_viridis(
    name = "Median Household Income",
    option = "plasma",
    trans = "log",  # Income is typically skewed, so log transform is helpful
    na.value = "grey90"
  ) +
  theme_minimal() +
  labs(
    title = "Median Household Income by Census Block Group (ACS 2019)",
    caption = "Data Source: U.S. Census Bureau, ACS 5-Year Estimates"
  )

print(ntl_bg_income_map)
ggsave(ntl_bg_income_map, filename = here::here("Outputs", "Figures", "ntl_bg_income_map.png"), 
       width = 10, height = 6, dpi = 300)
```

